---
title: "Plotting flood probability"
description: |
  Communicate flood probability in relatable terms.
author:
  - name: Michael Schramm
    url: https://michaelpaulschramm.com
    affiliation: Texas Water Resources Institute
    affiliation_url: https://twri.tamu.edu
date: "August 8, 2020"
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The public generally misunderstands the recurrence intervals used to describe flood events or storms. Typically, people believe the 100-year flood event as something that will only happen once every one hundred years. This is understandable given the description. However this recurrence interval means that there is a 1% chance of the event occurring in any given year. It might be better to explain the probability of a given flood event occurring over a time period. Say, 5 years or maybe over a 15 or 30 year mortgage. Then a person that moves into a 100-year floodplain would have some understanding of the probability of experiencing a flood event over the next n-years. We can use a binomial formula to calculate and graphically display these probabilities:

$$
P(k \text{ exceedances } n \text{ years})=\frac{n!}{k!(n-k)!}p^k(1-p)^{n-k}
$$

The relationship between probability $p$ and the recurrence interval $RI$, 
$RI=\frac{1}{p}$. In `R` we can utilize `dbinom()` to calculate the exceedance probability as: `dbinom(k, n, p)`.  


```{r data}
library(dplyr)
library(ggplot2)
library(hrbrthemes)

df <- tibble(ri = rep(c(50, 100, 500, 1000), each = 30), # return interval
             n = rep(1:30, 4), # n years
             aep = dbinom(1, n, 1/ri)) # exceedance probability
```

Now plot the probability of a flood event happening over n-years:

```{r plot, preview=TRUE}
ggplot(df) +
  geom_step(aes(n, aep, color = as.factor(ri))) +
  labs(x = "Number of years",
       y = "Probabililty of occurence") +
  scale_color_ft(name = "Flood Event", labels = c("50-yr flood",
                                                  "100-yr flood",
                                                  "500-yr flood",
                                                  "1000-yr flood")) +
  scale_x_continuous(expand = c(0, 0)) +
  theme_ft_rc() +
  theme(legend.position = "bottom",
        axis.title.y.right = element_blank())
```



