---
title: "Plotting drought conditions"
description: |
  Historic drought conditions have set in across much of the US.
author:
  - name: Michael Schramm
    url: https://michaelpaulschramm.com
    affiliation: Texas Water Resources Institute
    affiliation_url: https://twri.tamu.edu
date: "2022-07-22"
output:
  distill::distill_article:
    self_contained: false
base_url: https://michaelpaulschramm.com/
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      dev = "ragg_png",
                      fig.ext = "png",
                      dpi = 200)
library(prism)
library(fs)
library(terra)
library(tidyverse)
library(tidyterra)
library(scico)
library(mpsTemplates)
library(httr)
```

If you haven't been paying attention, the summer of 2022 is turning into a pretty historic drought event. Here in Texas we are seeing comparisons to 2011 and potentially worse.

<blockquote class="twitter-tweet"><p lang="en" dir="ltr">Austin-area reservoir levels compared to a long-term average, the previous two years, and 2011. Cool that the cool kids <a href="https://twitter.com/twdb?ref_src=twsrc%5Etfw">@twdb</a> added 2011 to their Water Data for Texas webtool: <a href="https://t.co/Tj3JCOqfVv">https://t.co/Tj3JCOqfVv</a> <a href="https://twitter.com/hashtag/txwater?src=hash&amp;ref_src=twsrc%5Etfw">#txwater</a> <a href="https://t.co/SiGz1NO1QW">pic.twitter.com/SiGz1NO1QW</a></p>&mdash; Robert Mace (@MaceatMeadows) <a href="https://twitter.com/MaceatMeadows/status/1550147326961786880?ref_src=twsrc%5Etfw">July 21, 2022</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>

<blockquote class="twitter-tweet"><p lang="en" dir="ltr">/2<br>2011: Heat built early but was steady. By July 21st, there were some triple-digit days but the worst of those were waiting for August &amp; September<br>2022: The extreme heat came quick &amp; fast. While the cooler year as a whole May - July&#39;s temps have been unprecedented for <a href="https://twitter.com/hashtag/bcstx?src=hash&amp;ref_src=twsrc%5Etfw">#bcstx</a> <a href="https://t.co/Pk1T6D5l70">pic.twitter.com/Pk1T6D5l70</a></p>&mdash; Shel Winkley (@KBTXShel) <a href="https://twitter.com/KBTXShel/status/1550293293795524609?ref_src=twsrc%5Etfw">July 22, 2022</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>


There are a lot of great datasets to explore for temperature, precipitation, and drought. Here are some R packages that I will use:

```{r echo=TRUE, eval=FALSE}
library(prism)
library(terra)
library(tidyverse)
library(tidyterra)
library(scico)
library(httr)

# I use the mpsTemplates package to streamline my ggplot, just sub theme_minimal
# or similar.
```


## Precipitation

The [`prism`](https://docs.ropensci.org/prism/index.html) 📦 provides streamlined access to the [Oregon State University PRISM Climate Group](https://www.prism.oregonstate.edu/) data. This includes, modeled 30-year gridded normal precipitation, temperature and other data over the contiguous US. Data is available at 4 km and 800 m resolutions. The 30-year normals for April, May and June are shown below.
 
```{r prismsetting, message=FALSE, warning=FALSE, include=FALSE}
data_path <- fs::path_wd("data")
prism::prism_set_dl_dir(data_path)

```

```{r normals, echo=TRUE, fig.height=8, fig.width=8, message=FALSE, warning=FALSE, results="hide", fig.keep="all", out.width="100%"}
## download monthly normals
get_prism_normals(type = "ppt",
                  resolution = "4km",
                  mon = 1:6,
                  keepZip = FALSE)

## return the folders
ppt_norm_1 <- prism_archive_subset(type = "ppt",
                                   temp_period = "monthly normals",
                                   mon = 1:6,
                                   resolution = "4km")

## convert to terra
ppt_norm_1 <- pd_to_file(ppt_norm_1)
ppt_norm_1 <- rast(ppt_norm_1)


## change the layer names to something sensible
names(ppt_norm_1) <- month.name[1:6]

## plot
ggplot() +
  geom_spatraster(data = ppt_norm_1) +
  scale_fill_scico(name = "Precipitation (mm)", palette = "lapaz", direction = -1,
                   trans = "pseudo_log",
                   breaks = c(0,10,100,500),
                   na.value = "transparent") +
  coord_sf(crs = 2163) +
  facet_wrap(~lyr, ncol = 2, nrow = 3) +
  theme_mps_noto() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        strip.text = element_text(hjust = 0, size = 10),
        strip.background = element_blank(),
        legend.position = "right",
        legend.direction = "vertical",
        legend.key.height = ggplot2::unit(40L, "pt")) +
  labs(caption = "1991-2020 (30-year) Monthly Normal Precipitation\nSource: PRISM Climate Group (https://www.prism.oregonstate.edu/)")

```


We can also obtain gridded total monthly precipitation for each month of 2022 (so far) and calculate the precipitation anomaly for each month. The function to download monthly precipitation is shown, the code to plot monthly data is the same as above and isn't shown. It is immediately evident the January through May 2022 look very rough for the southwest US compared to the average monthly precipitation totals.

```{r monthly, echo=TRUE, message=FALSE, warning=FALSE, results="hide"}
## download monthly data
get_prism_monthlys(type = "ppt",
                   year = 2022,
                   mon = 1:6,
                   keepZip = FALSE)
```


```{r monthlyfig, echo=FALSE, fig.height=8, fig.width=8, message=FALSE, warning=FALSE, results="hide", fig.keep="all", out.width="100%"}
ppt_1 <- prism_archive_subset(type = "ppt",
                              temp_period = "monthly",
                              mon = 1:6)


## convert to terra
ppt_1 <- pd_to_file(ppt_1)
ppt_1 <- rast(ppt_1)

## change the layer names to something sensible
names(ppt_1) <- month.name[1:6]

## plot
ggplot() +
  geom_spatraster(data = ppt_1) +
  scale_fill_scico(name = "Precipitation (mm)", palette = "lapaz", direction = -1,
                   trans = "pseudo_log",
                   breaks = c(0,10,100,200,400),
                   na.value = "transparent") +
  coord_sf(crs = 2163) +
  facet_wrap(~lyr, ncol = 2, nrow = 3) +
  theme_mps_noto() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        strip.text = element_text(hjust = 0, size = 10),
        strip.background = element_blank(),
        legend.position = "right",
        legend.direction = "vertical",
        legend.key.height = ggplot2::unit(40L, "pt")) +
  labs(caption = "2022 Total Monthly Precipitation\nSource: PRISM Climate Group (https://www.prism.oregonstate.edu/)")
```

The precipitation anomaly for a given month can be quickly calculated and plotted. We can use the `terra` 📦 to calculate the percentage difference from the 30-year monthly mean precipitation.

```{r anomaly, echo=TRUE, fig.height=8, fig.width=8, message=FALSE, warning=FALSE, results="hide", fig.keep="all", out.width="100%"}
anomaly <- ((ppt_1 - ppt_norm_1)/ppt_norm_1)*100

ggplot() +
  geom_spatraster(data = anomaly) +
  scale_fill_scico(name = "Precipitation Anomaly\n(% difference)", 
                   palette = "vik", 
                   direction = -1,
                   breaks = c(-100,0,100,200,300,400),
                   limits = c(-100,400),
                   ## some infinite and very high values, will squish them to 400
                   oob = scales::oob_squish_any,
                   ## this centers us on zero
                   values = scales::rescale(c(-100,0,400)),
                   na.value = "transparent") +
  coord_sf(crs = 2163) +
  facet_wrap(~lyr, ncol = 2, nrow = 3) +
  theme_mps_noto() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        strip.text = element_text(hjust = 0, size = 10),
        strip.background = element_blank(),
        legend.position = "right",
        legend.direction = "vertical",
        legend.key.height = ggplot2::unit(40L, "pt")) +
  labs(caption = "2022 Monthly Precipitation Anomalies\nSource: PRISM Climate Group (https://www.prism.oregonstate.edu/)")
  
```

## Temperature

PRISM also provides minimum, maximum, and mean daily temperature data. Anomalies for the year are shown below.

```{r tanomaly, echo=FALSE, eval=TRUE, fig.height=8, fig.width=8, message=FALSE, warning=FALSE, results="hide", fig.keep="all", out.width="100%"}
## download monthly normals
get_prism_normals(type = "tmean",
                  resolution = "4km",
                  mon = 1:6,
                  keepZip = FALSE)
## download monthly data
get_prism_monthlys(type = "tmean",
                   year = 2022,
                   mon = 1:6,
                   keepZip = FALSE)
## return the folders
tmean_norm_1 <- prism_archive_subset(type = "tmean",
                                     temp_period = "monthly normals",
                                     mon = 1:6,
                                     resolution = "4km")
tmean_1 <- prism_archive_subset(type = "tmean",
                                temp_period = "monthly",
                                mon = 1:6)

## convert to terra
tmean_norm_1 <- pd_to_file(tmean_norm_1)
tmean_norm_1 <- rast(tmean_norm_1)

tmean_1 <- pd_to_file(tmean_1)
tmean_1 <- rast(tmean_1)

## change the layer names to something sensible
names(tmean_norm_1) <- month.name[1:6]
names(tmean_1) <- month.name[1:6]

tanomaly <- tmean_1 - tmean_norm_1

ggplot() +
  geom_spatraster(data = tanomaly) +
  scale_fill_scico(name = "Temperature\nAnomaly (°C)", 
                   palette = "vik", midpoint = 0,
                   na.value = "transparent") +
  coord_sf(crs = 2163) +
  facet_wrap(~lyr, ncol = 2, nrow = 3) +
  theme_mps_noto() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        strip.text = element_text(hjust = 0, size = 10),
        strip.background = element_blank(),
        legend.position = "right",
        legend.direction = "vertical",
        legend.key.height = ggplot2::unit(40L, "pt")) +
  labs(caption = "2022 Daily Mean Temperature Anomalies\nSource: PRISM Climate Group (https://www.prism.oregonstate.edu/)")
```

## Drought

The [US Drought Monitor](https://www.drought.gov/data-maps-tools/us-drought-monitor) provides weekly maps that categorize the locations and intensity of drought across the US. The National Drought Monitoring Center provides some [REST services](https://droughtmonitor.unl.edu/DmData/DataDownload/WebServiceInfo.aspx) to programatically download data. This example will download and plot area statistics for drought categories across Texas.

```{r echo=TRUE, results=FALSE}
## download data using the REST API. More info here:
## https://droughtmonitor.unl.edu/DmData/DataDownload/WebServiceInfo.aspx

url <- "https://usdmdataservices.unl.edu/api/StateStatistics/GetDroughtSeverityStatisticsByAreaPercent?aoi=48&startdate=1/1/2000&enddate=7/15/2022&statisticsType=1"
r <- GET(url, accept("text/csv"))
df <- content(r)
```

```{r txdrought, echo=TRUE, eval=TRUE, fig.height=5, fig.width=8, message=FALSE, warning=FALSE, results="hide", fig.keep="all", out.width="100%"}
df |> 
  ## the values need to be adjusted because category D0
  ## represents D0-D4, D1 is D1-D4, etc.
  mutate(D0 = D0-D1,
         D1 = D1-D2,
         D2 = D2-D3,
         D3 = D3-D4) |> 
  pivot_longer(cols = c("D0", "D1", "D2", "D3", "D4"),
               names_to = "Drought_Category") |> 
  mutate(Drought_Category = as_factor(Drought_Category)) |>
  mutate(Drought_Cateogry = fct_relevel(Drought_Category, "D0", "D1", "D2", "D3", "D4")) |> 
  ggplot() +
  geom_col(aes(ValidStart, value, fill = Drought_Category)) +
  scale_fill_scico_d(palette = "lajolla", name = "USDM Category")  +
  scale_y_continuous(expand = c(0,0)) +
  scale_x_date(date_breaks = "2 years",
               date_labels = "%Y",
               expand = c(0,0)) +
  theme_mps_noto() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        strip.text = element_text(hjust = 0, size = 10),
        strip.background = element_blank(),
        legend.position = "right",
        legend.direction = "vertical",
        legend.key.height = ggplot2::unit(40L, "pt")) +
  labs(caption = "USDM Drought Categories in Texas by Percent Area\nSource: U.S. Drought Monitor (https://droughtmonitor.unl.edu/Data.aspx)",
       x = "Date", y = "Percent Area")
```

