\documentclass[11pt,]{article}
\usepackage[sc, osf]{mathpazo}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}




\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

% Now begins the stuff that I added.
% ----------------------------------

% Custom section fonts
\usepackage{sectsty}
\sectionfont{\rmfamily\mdseries\large\bf}
\subsectionfont{\rmfamily\mdseries\normalsize\itshape}


% Make lists without bullets
\renewenvironment{itemize}{
  \begin{list}{}{
    \setlength{\leftmargin}{1.5em}
  }
}{
  \end{list}
}


% Make parskips rather than indent with lists.
\usepackage{parskip}
\usepackage{titlesec}
\titlespacing\section{0pt}{12pt plus 4pt minus 2pt}{4pt plus 2pt minus 2pt}
\titlespacing\subsection{0pt}{12pt plus 4pt minus 2pt}{4pt plus 2pt minus 2pt}

% Use fontawesome. Note: you'll need TeXLive 2015. Update.
\usepackage{fontawesome}

% Fancyhdr, as I tend to do with these personal documents.
\usepackage{fancyhdr,lastpage}
\pagestyle{fancy}
\renewcommand{\headrulewidth}{0.0pt}
\renewcommand{\footrulewidth}{0.0pt}
\lhead{}
\chead{}
\rhead{}
\lfoot{
\cfoot{\scriptsize  William Sealy Gosset - CV }}
\rfoot{\scriptsize \thepage/{\hypersetup{linkcolor=black}\pageref{LastPage}}}

% Always load hyperref last.
\usepackage{hyperref}
\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref

\hypersetup{unicode=true,
            pdftitle={William Sealy Gosset:  CV (Curriculum Vitae)},
            pdfauthor={William Sealy Gosset},
            pdfkeywords={R Markdown, academic CV, template},
            colorlinks=true,
            linkcolor=blue,
            citecolor=Blue,
            urlcolor=blue,
            breaklinks=true, bookmarks=true}
\urlstyle{same}  % don't use monospace font for urls

\begin{document}


\centerline{\huge \bf William Sealy Gosset}

\vspace{2 mm}

\hrule

\vspace{2 mm}

\moveleft.5\hoffset\centerline{Chief Brewer, Arthur Guinness \& Son}
\moveleft.5\hoffset\centerline{Guinness Brewery · Park Royal · London NW10 7RR, UK}
\moveleft.5\hoffset\centerline{ \faEnvelopeO \hspace{1 mm} \href{mailto:}{\tt \href{mailto:guinness@consumer-care.net}{\nolinkurl{guinness@consumer-care.net}}} \hspace{1 mm}  \faPhone \hspace{1 mm}  +353 1 408 4800  \hspace{1 mm}     \faGlobe \hspace{1 mm} \href{http://guinness.com}{\tt guinness.com}   }

\vspace{2 mm}

\hrule


\section{Employment}\label{employment}

\emph{Arthur Guinness \& Son}

Chief Brewer (Park Royal) \hfill 1935-1937

Researcher (Dublin) \hfill 1899-1935

\section{Education}\label{education}

\emph{New College, Oxford University}, B.A. Chemistry \hfill 1899

\section{Publications}\label{publications}

\subsection{Books}\label{books}

Pearson, E.S. and John Wishart (eds.). 1943. \emph{``Student's''
Collected Papers} London: Biometrika Office.

\subsection{Journal Articles}\label{journal-articles}

Student. 1938. ``Comparison between Balanced and Random Arrangements on
Field Plots.'' \emph{Biometrika} 29(3-4): 363--378.

Student. 1936. ``Co-operation in Large Scale Experiments.''
\emph{Supplement to Journal of the Royal Statistical Society} 3(2):
115--136.

Student. 1931. ``On the \emph{z} Test.'' \emph{Biometrika} 23(3-4):
407-408.

Student. 1931. ``The Lanarkshire Milk Experiment.'' \emph{Biometrika}
23(3-4): 398--406.

Student. 1929. ``Statistics in Biological Research.'' \emph{Nature} 124:
93

Student. 1927. ``Errors of Routine Analysis.'' \emph{Biometrika}
19(1-2): 151--164.

Student. 1926. ``Mathematics and Agronomy.'' \emph{Journal of the
American Society of Agronomy} 18.

Student. 1925. ``New Tables for Testing the Significance of
Observations.'' \emph{Metron} 5(3): 105--108.

Student. 1923. ``On Testing Varieties of Cereals.'' \emph{Biometrika}
15(3-4): 271--293.

Student. 1921. ``An Experimental Determination of the Probable Error of
Dr.~Spearman's Correlation Co-efficient'' \emph{Biometrika} 13(2-3):
263--282.

Student. 1917. ``Tables for Estimating the Probability That the Given
Mean of a Unique Sample of Observations Lies between \(-\infty\) and Any
Given Distance of the Mean of the Population from Which the Sample Is
Drawn.'' \emph{Biometrika} 11(4): 414--417.

Student. 1909. ``The Distribution of the Means of Samples Which Are Not
Drawn at Random.'' \emph{Biometrika} 7(1-2): 210--214.

Student. 1908. ``Probable Error of a Correlation Coefficient.''
\emph{Biometrika} 6(2-3): 302--310.

Student. 1908. ``The Probable Error of a Mean.'' \emph{Biometrika} 6(1):
1--25.

Student. 1907. ``On the Error of Counting with a Haemacytometer.''
\emph{Biometrika} 5(3): 351--360.

\subsection{Other Publications}\label{other-publications}

Student. 1905. ``The Pearson Co-efficient of Correlation.'' Supplement
to Report of 1904, Arthur Guinness Son \& Company.

Student. 1904. ``The Application of the `Law of Error' to the Work of
the Brewery.'' Report, Arthur Guinness Son \& Company.

\section{References}\label{references}

Available upon request, though likely would include Karl Pearson and
Ronald A. Fisher. Please send corrections, feedback, or details of
missed publications (of which there are probably a few) to
\href{mailto:svmille@clemson.edu}{\nolinkurl{svmille@clemson.edu}}.

\end{document}